---
title: "subs"
output: pdf_document
date: "2025-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in nflverse data

We will take the 2024 data

```{r}
library(nflreadr)
library(dplyr)

# Load the datasets
pbp <- load_pbp(seasons = 2023)
participation <- load_participation(seasons = 2023)

# Merge with correct column mapping
merged_data <- pbp %>%
  left_join(
    participation,
    by = c("game_id" = "nflverse_game_id", "play_id" = "play_id")
  )
```


in merged_data grab play_id, game_id, home_team, away_team, season_type, week, offense_players and defense_players


```{r}
merged_data <- merged_data %>%
  select(play_id, game_id, desc, home_team, away_team, posteam, down, play_type, season_type, week, offense_players, defense_players)

# use home and away to create def team... then kill home and away columns
merged_data$def_team <- ifelse(merged_data$posteam == merged_data$home_team, merged_data$away_team, merged_data$home_team)
merged_data <- merged_data %>%
  select(-home_team, -away_team)

# keep only if play type is run or pass and kill any rows that have NA for down
merged_data <- merged_data %>%
  filter(play_type %in% c("run", "pass")) %>%
  filter(!is.na(down))
```

```{r}
str(merged_data)
```

```{r}
library(tidyr)
library(dplyr)

# Split offense players (up to 11 positions)
merged_data <- merged_data %>%
  separate(
    offense_players,
    into = paste0("off_player_", 1:11),
    sep = ";",
    remove = FALSE,  # Keep original column
    fill = "right",  # Fill with NA if fewer than 11
    convert = TRUE   # Convert to appropriate data type
  ) %>%
  # Split defense players (up to 11 positions)
  separate(
    defense_players,
    into = paste0("def_player_", 1:11),
    sep = ";",
    remove = FALSE,
    fill = "right",
    convert = TRUE
  )
# kill the od cols now
merged_data <- merged_data %>%
  select(-offense_players, -defense_players)

# Verify the new structure
glimpse(merged_data)

```

Now we need to go in each game drive by drive... basicallly just count between plays on the same drives how many players on offense/defense swapped out

To identify a game its the game id.. that is easy

for the drives it can be sequential!


 so go for each game... then take first value as down 1 then 2 then back down to 1 mabye or whatever but a change in drive is when the posteam switches... when it switches you start recounting.... the output need to be by game then by drive and just make a drive index... then count the number of players that change between plays... so you are looking for each player on offense how many swapped and how many swapped on defense... it does not need to be slot specific just them being on the field is fine.

```{r}
# Create a function to identify drives by team possession changes
identify_drives <- function(data) {
  # Sort by game_id and play_id to ensure chronological ordering
  data <- data %>%
    arrange(game_id, play_id)
  
  # Initialize drive variables
  data$drive_number <- 0
  current_drive <- 1
  current_game <- ""
  current_posteam <- ""
  
  # Identify drives by detecting changes in possession team
  for (i in 1:nrow(data)) {
    if (data$game_id[i] != current_game) {
      # New game
      current_game <- data$game_id[i]
      current_posteam <- data$posteam[i]
      current_drive <- 1
    } else if (data$posteam[i] != current_posteam) {
      # New drive (possession team changed)
      current_posteam <- data$posteam[i]
      current_drive <- current_drive + 1
    }
    
    # Assign drive number
    data$drive_number[i] <- current_drive
  }
  
  return(data)
}

# Function to count player substitutions between consecutive plays
# Function to count player substitutions correctly
count_substitutions <- function(data) {
  # Create a unique identifier for each drive
  data$drive_id <- paste(data$game_id, data$drive_number, sep = "_")
  
  # Initialize results dataframe
  substitution_data <- data.frame()
  
  # Process each drive
  for (current_drive_id in unique(data$drive_id)) {
    # Get plays for this drive, ordered by play_id
    drive_plays <- data %>%
      filter(drive_id == current_drive_id) %>%
      arrange(play_id)
    
    # Skip if only one play in the drive
    if (nrow(drive_plays) <= 1) {
      next
    }
    
    # Compare consecutive plays
    for (i in 1:(nrow(drive_plays) - 1)) {
      current_play <- drive_plays[i, ]
      next_play <- drive_plays[i + 1, ]
      
      # Extract offensive players for current and next play
      current_off_players <- unlist(current_play[grep("off_player_", names(current_play))])
      current_off_players <- current_off_players[!is.na(current_off_players)]
      
      next_off_players <- unlist(next_play[grep("off_player_", names(next_play))])
      next_off_players <- next_off_players[!is.na(next_off_players)]
      
      # Extract defensive players for current and next play
      current_def_players <- unlist(current_play[grep("def_player_", names(current_play))])
      current_def_players <- current_def_players[!is.na(current_def_players)]
      
      next_def_players <- unlist(next_play[grep("def_player_", names(next_play))])
      next_def_players <- next_def_players[!is.na(next_def_players)]
      
      # CORRECT COUNT: How many players changed (not double counting)
      # For offense: Count players who left (or entered, should be same number)
      off_left <- setdiff(current_off_players, next_off_players)
      off_subs <- length(off_left)
      
      # For defense: Count players who left
      def_left <- setdiff(current_def_players, next_def_players) 
      def_subs <- length(def_left)
      
      # Create row for this play transition
      result_row <- data.frame(
        game_id = current_play$game_id,
        week = current_play$week,
        season_type = current_play$season_type,
        drive_id = current_drive_id,
        drive_number = current_play$drive_number,
        play_id = current_play$play_id,
        next_play_id = next_play$play_id,
        posteam = current_play$posteam,
        def_team = current_play$def_team,
        play_type = current_play$play_type,
        down = current_play$down,
        off_subs = off_subs,
        def_subs = def_subs,
        stringsAsFactors = FALSE
      )
      
      # Add to results
      substitution_data <- rbind(substitution_data, result_row)
    }
  }
  
  return(substitution_data)
}

# Create simple summary matrices for plotting
create_summary_matrices <- function(substitution_data) {
  # 1. Team level summary (offensive substitutions)
  team_off_summary <- substitution_data %>%
    group_by(posteam) %>%
    summarize(
      total_plays = n(),
      total_off_subs = sum(off_subs),
      avg_off_subs_per_play = mean(off_subs),
      .groups = 'drop'
    ) %>%
    arrange(desc(avg_off_subs_per_play))
  
  # 2. Team level summary (defensive substitutions)
  team_def_summary <- substitution_data %>%
    group_by(def_team) %>%
    summarize(
      total_plays = n(),
      total_def_subs = sum(def_subs),
      avg_def_subs_per_play = mean(def_subs),
      .groups = 'drop'
    ) %>%
    arrange(desc(avg_def_subs_per_play))
  
  # 3. Down level summary
  down_summary <- substitution_data %>%
    group_by(down) %>%
    summarize(
      total_plays = n(),
      avg_off_subs = mean(off_subs),
      avg_def_subs = mean(def_subs),
      .groups = 'drop'
    )
  
  # 4. Play type summary
  play_type_summary <- substitution_data %>%
    group_by(play_type) %>%
    summarize(
      total_plays = n(),
      avg_off_subs = mean(off_subs),
      avg_def_subs = mean(def_subs),
      .groups = 'drop'
    )
  
  # 5. Game level summary
  game_summary <- substitution_data %>%
    group_by(game_id) %>%
    summarize(
      total_plays = n(),
      total_off_subs = sum(off_subs),
      total_def_subs = sum(def_subs),
      avg_off_subs = mean(off_subs),
      avg_def_subs = mean(def_subs),
      .groups = 'drop'
    )
  
  # 6. Week level summary
  week_summary <- substitution_data %>%
    group_by(week) %>%
    summarize(
      total_plays = n(),
      avg_off_subs = mean(off_subs),
      avg_def_subs = mean(def_subs),
      .groups = 'drop'
    )
  
  # Return all summaries in a list
  return(list(
    team_offense = team_off_summary,
    team_defense = team_def_summary,
    by_down = down_summary,
    by_play_type = play_type_summary,
    by_game = game_summary,
    by_week = week_summary,
    raw_data = substitution_data
  ))
}

# Apply the functions to the data
# Step 1: Identify drives
message("Identifying drives...")
merged_data_with_drives <- identify_drives(merged_data)

# Step 2: Count substitutions between consecutive plays
message("Counting substitutions between plays...")
substitution_counts <- count_substitutions(merged_data_with_drives)

# Step 3: Create summary matrices for plotting
message("Creating summary matrices...")
substitution_summaries <- create_summary_matrices(substitution_counts)

```

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Assuming substitution_counts is already loaded
# If running this as a separate script, load the data:
# substitution_counts <- read.csv("substitution_raw_data.csv")

# Create count tables for each value
off_counts <- table(substitution_counts$off_subs)
def_counts <- table(substitution_counts$def_subs)

# Convert to data frames for plotting
off_df <- data.frame(
  subs = as.numeric(names(off_counts)),
  count = as.numeric(off_counts)
)

def_df <- data.frame(
  subs = as.numeric(names(def_counts)),
  count = as.numeric(def_counts)
)

# Only keep values 0-10 for cleaner visualization
off_df <- off_df[off_df$subs <= 10, ]
def_df <- def_df[def_df$subs <= 10, ]

# 1. Offensive substitution histogram with counts
off_hist <- ggplot(off_df, aes(x = factor(subs), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black", alpha = 0.7) +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  labs(
    title = "Frequency of Offensive Substitutions per Play",
    x = "Number of Offensive Substitutions",
    y = "Frequency"
  ) +
  theme_minimal() +
  scale_x_discrete(limits = as.character(0:10))

# 2. Defensive substitution histogram with counts
def_hist <- ggplot(def_df, aes(x = factor(subs), y = count)) +
  geom_bar(stat = "identity", fill = "firebrick", color = "black", alpha = 0.7) +
  geom_text(aes(label = count), vjust = -0.5, size = 3) +
  labs(
    title = "Frequency of Defensive Substitutions per Play",
    x = "Number of Defensive Substitutions",
    y = "Frequency"
  ) +
  theme_minimal() +
  scale_x_discrete(limits = as.character(0:10))

# 3. Combined substitution counts by down
# First, prepare data
down_subs <- substitution_counts %>%
  filter(off_subs <= 10, def_subs <= 10) %>% # Keep reasonable values
  group_by(down, off_subs) %>%
  summarize(count = n(), .groups = 'drop')

# Create stacked bar chart for offensive subs by down
down_off_hist <- ggplot(down_subs, aes(x = factor(off_subs), y = count, fill = factor(down))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = count), position = position_dodge(width = 0.9), vjust = -0.5, size = 2.5) +
  labs(
    title = "Offensive Substitutions by Down",
    x = "Number of Substitutions",
    y = "Frequency",
    fill = "Down"
  ) +
  theme_minimal() +
  scale_x_discrete(limits = as.character(0:10))

# 4. Substitutions by play type
play_type_subs <- substitution_counts %>%
  filter(off_subs <= 10) %>% # Keep reasonable values
  group_by(play_type, off_subs) %>%
  summarize(count = n(), .groups = 'drop')

play_type_hist <- ggplot(play_type_subs, aes(x = factor(off_subs), y = count, fill = play_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = count), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(
    title = "Offensive Substitutions by Play Type",
    x = "Number of Substitutions",
    y = "Frequency",
    fill = "Play Type"
  ) +
  theme_minimal() +
  scale_x_discrete(limits = as.character(0:10))

# Display plots
print(off_hist)
print(def_hist)
print(down_off_hist)
print(play_type_hist)

# Create a simple summary stats table
summary_stats <- data.frame(
  Statistic = c(
    "Mean Offensive Substitutions",
    "Median Offensive Substitutions",
    "Most Common Offensive Substitution",
    "Mean Defensive Substitutions",
    "Median Defensive Substitutions",
    "Most Common Defensive Substitution"
  ),
  Value = c(
    round(mean(substitution_counts$off_subs), 2),
    median(substitution_counts$off_subs),
    as.numeric(names(off_counts)[which.max(off_counts)]),
    round(mean(substitution_counts$def_subs), 2),
    median(substitution_counts$def_subs),
    as.numeric(names(def_counts)[which.max(def_counts)])
  )
)

print(summary_stats)
```


```{r}
# Calculate average substitutions by down for both offense and defense
avg_subs_by_down <- substitution_counts %>%
  group_by(down) %>%
  summarize(
    avg_off_subs = mean(off_subs, na.rm = TRUE),
    avg_def_subs = mean(def_subs, na.rm = TRUE),
    count = n(),
    .groups = 'drop'
  )

# Print the table
print(avg_subs_by_down)

# Convert to long format for easier plotting
avg_subs_long <- avg_subs_by_down %>%
  pivot_longer(
    cols = c(avg_off_subs, avg_def_subs),
    names_to = "team_type",
    values_to = "avg_subs"
  ) %>%
  mutate(
    team_type = ifelse(team_type == "avg_off_subs", "Offense", "Defense")
  )

# Create bar plot
ggplot(avg_subs_long, aes(x = factor(down), y = avg_subs, fill = team_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.2f", avg_subs)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, 
            size = 3.5) +
  labs(
    title = "Average Substitutions by Down",
    subtitle = paste("Based on", sum(avg_subs_by_down$count), "plays"),
    x = "Down",
    y = "Average Number of Substitutions",
    fill = "Team Type"
  ) +
  scale_fill_manual(values = c("Offense" = "steelblue", "Defense" = "firebrick")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.title = element_text(face = "bold")
  )

# Create a version with error bars (standard error)
avg_subs_with_se <- substitution_counts %>%
  group_by(down) %>%
  summarize(
    avg_off_subs = mean(off_subs, na.rm = TRUE),
    se_off_subs = sd(off_subs, na.rm = TRUE) / sqrt(n()),
    avg_def_subs = mean(def_subs, na.rm = TRUE),
    se_def_subs = sd(def_subs, na.rm = TRUE) / sqrt(n()),
    count = n(),
    .groups = 'drop'
  )

# Convert to long format for error bar plotting
avg_subs_se_long <- avg_subs_with_se %>%
  pivot_longer(
    cols = c(avg_off_subs, avg_def_subs),
    names_to = "team_type",
    values_to = "avg_subs"
  ) %>%
  mutate(
    se = ifelse(team_type == "avg_off_subs", se_off_subs, se_def_subs),
    team_type = ifelse(team_type == "avg_off_subs", "Offense", "Defense")
  )

# Create bar plot with error bars
ggplot(avg_subs_se_long, aes(x = factor(down), y = avg_subs, fill = team_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(ymin = avg_subs - se, ymax = avg_subs + se),
    position = position_dodge(width = 0.9),
    width = 0.25
  ) +
  geom_text(aes(label = sprintf("%.2f", avg_subs)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 3.5) +
  labs(
    title = "Average Substitutions by Down (with Standard Error)",
    subtitle = paste("Based on", sum(avg_subs_with_se$count), "plays"),
    x = "Down",
    y = "Average Number of Substitutions",
    fill = "Team Type"
  ) +
  scale_fill_manual(values = c("Offense" = "steelblue", "Defense" = "firebrick")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.title = element_text(face = "bold")
  )

# Table with percentages compared to average
overall_avg <- substitution_counts %>%
  summarize(
    avg_off_subs = mean(off_subs, na.rm = TRUE),
    avg_def_subs = mean(def_subs, na.rm = TRUE)
  )

comparison_table <- avg_subs_by_down %>%
  mutate(
    off_pct_diff = (avg_off_subs / overall_avg$avg_off_subs - 1) * 100,
    def_pct_diff = (avg_def_subs / overall_avg$avg_def_subs - 1) * 100
  ) %>%
  select(down, avg_off_subs, off_pct_diff, avg_def_subs, def_pct_diff, count)

# Print comparison table
print(comparison_table)
```



